TODO: title

### 2014-04-25 paper/main.tex
done working for the day

\section{Methodology}
  % (high level description/restatement of approach) To answer our research questions we will various metrics of recommenders in an offline setting simulations users as they join the system.
    % maybe hit some limitations
    % motiviate _why do a simulation_ (a lot of metrics we want to try)
  % temporal replay evaluation 
    % similar technique, harder to do than ours, and with a different goal.
  % SOMEWHERE AROUND HERE talk about drnner and adaptive bootstrapping and 10 isenough paper's ad hoc plots by datapoint size.
  % algorithms (discuss tuning of algorithms per algorithm?)
    % item mean baseline
    % user adjusted baseline (@michael, is there a cite for this?)
    % user-user
    % item-item
    % SVD
  % metrics (expected / good behavior for each metric (I.E. this should go down, any reasonable algorithm should always beet baseline)) 
    % accuracy metrics 
      % RMSE
    % recommendation performance metrics (list subject to change)
      % ndcg
      % MAP@N (not yet measured)
      % precision@N
      % fallout @ N
      % RMSE of recommended items
      % (note that we looked at MRR and recall, but found them to behave the same).
    % other properties
      % recommendation average popularity
      % recommendation diversity
      % spread (entropy of recommendations) (related to recommendation coverage)

  % dataset
    % # user
    % # items
    % distributions of # ratings per user (important: min 20)
  % subsampling strategy
    % describe normal crossfold solution
    % describe issues with solution
    % describe our crossfold then subsample solution.

@@

  Using the methodology described in \cite{sec:methodology} we compute a value for each metric on each of the five algorithms.
  We plot these results against the number of items retained.
  These plots allow us to compare the algorithms against each other, and also, to understand how each algorithm's performance changes with more data.
  Each plot will have error bars showing the 95\% confidence intervals around the average per use value for the metric.

\subsection{Prediction accuracy}
  % Prediction accuracy
    % RMSE plot

TODO: rmse plot

    % point out error bars etc.
    % explain / interpret plot
  Figure \ref{fig:rmse} shows how accurate our algorithms are as measured by RMSE.
  As would be expected the accuracy of our item baseline doesn't change as the amount of information changes.
  The User-User and svd algorithms both perform as expected, starting out with the same accuracy as the item baseline, and then monotonically improving.
  User-User does show a small increase of error with only one rating, but the change is quite small and unlikely to be noticed by users.
  Interestingly, User-User and svd both perform about as well as our smarter user item baseline. %TODO: explain this better

@@

@@

  % TopN average rating / topn rmse
    % interpret graph and reason about what it means about quality
  The first graph in figure \ref{fig:topN.rmse} shows the average number of items in the top 20 recommendations for each algorithm that were also in the test set.
  The average number of test set items in recommendations shows a similar trend to the MAP, precision, and recall metrics, user-user getting by far the fewest seen items, and the baseline getting the most.
  %TODO: talk about trust building here!

@@

\subsection{Other recommender metrics}
  % Other recommender metrics
    % popularity graph
  The first graph in figure \ref{fig:pop} shows the average popularity of recommended items by each algorithm.
  %TODO: re-write this
  Across all datapoint sizes the ranking of algorithms is the same, with the baselines recommending the most popular movies, followed by svd, which recommends less popular movies, then Item-Item, which recommends much less popular movies, leaving User-User to recommend the least popular movies.
  Unfortunately, we don't know of any guideline for understanding how popular users desire their recommendations to be.
  That in mind, we expect that users will feel the baseline's popularity (around 1000) is too high, and User-User's popularity (around 50) is too low.
  SVD and Item-Item both seem to perform well, with SVD generating more popular recommendations, than Item-Item, and both algorithms showing a significant trend to make less popular predictions as we know more about the user.

@@

>>>>

\section{Methodology}
\label{sec:methodology}
  % (high level description/restatement of approach) To answer our research questions we will various metrics of recommenders in an offline setting simulations users as they join the system.
    % maybe hit some limitations
    % motiviate _why do a simulation_ (a lot of metrics we want to try)

  To answer our research questions we will evaluate algorithms in an offline simulation designed to simulate users as they join the system.
  We will do this by running separate instances of the standard train and test methodology, each one designed so that test users have some small number of ratings.
  We will then plot metric results against the number of retained items to understand both how the algorithms compare on these metrics, and how the algorithm performance changes with the number of ratings provided.
  This, or a similar technique, has been used by previous authors \cite{DrennerInitialExperiance, TenIsEnough, AdaptiveBootstrap}, however, these normally are used with only one algorithm or metric, and normally just to make the point that prediction accuracy improves as the number of ratings grows.
  We will use this technique to develop a much more nuanced understanding of the differences between algorithms, and how they behave for users with few ratings.

  \subsection*{Algorithms}
  % algorithms (discuss tuning of algorithms per algorithm?)
  % motivation for algorithm selection
  % list algorithms

  We seek to develop an understanding of how a range of different standard algorithms perform at for new users.
  Rather than picking advanced algorithms that are tuned for new user performance, we decided to first understand how representatives of three common types of algorithms perform.
  Therefore we will compare User-User, Item-Item, and Funk's SVD algorithms against two simple baseline algorithms.
  More information about these algorithms can be found in table \ref{tbl:algo}
  Code for our evaluation, using the Lenskit evaluation framework \cite{lenskit} is available at \url{https://bitbucket.org/kluver/coldstartrecommendation}.

  \begin{table}
    \centering
    \begin{tabular}{|p{6em}|p{18em}|}
      \hline
      Algorithm        & Algorithm description                                                                                     \\\hline
      User-User        & User based nearest neighbor collaborative filtering \cite{resnick1994grouplens}                           \\\hline
      Item-Item        & Item based nearest neighbor collaborative filtering \cite{sarwar2001item}                                 \\\hline
      SVD              & Simon Funk's SVD based collaborative filtering approach \cite{funk_netflix_2006}                          \\\hline
      ItemBaseline     & Item's average rating, slight mean damping is applied                                                     \\\hline
      UserItem\-Baseline & ItemBasline adjusted by the user's average offset from the ItemBaseline. slight mean damping is applied \\\hline
    \end{tabular}
    \caption{Summary of algorithms used in our research.}
    \label{tbl:algo}
  \end{table}

  % evaluation and code to run these algortihms available at.
  % citations and lenskit info
  %need new topic sentence.

  Each algorithm was tuned using the values from Ekstrand's 2012 lenskit evaluation \cite{ekstrand2012recommenders}.
  Sine this evaluation was on a slightly different dataset (MovieLens 10M), we performed only minor tuning from these parameters.
  Most notably we found the tuning of the damping parameter used in the baseline recommenders to be too large, which limited accuracy at cold start.
  Overall, we did not find our qualitative results to be highly sensitive to minor tuning changes, although specific values may change slightly.

  \subsection*{Metrics}

  % motivate many metrics.
  % metrics (expected / good behavior for each metric (I.E. this should go down, any reasonable algorithm should always beet baseline)) 
    % accuracy metrics 
      % RMSE
    % recommendation performance metrics (list subject to change)
      % ndcg
      % MAP@N (not yet measured)
      % precision@N
      % fallout @ N
      % RMSE of recommended items
      % (note that we looked at MRR and recall, but found them to behave the same).
    % other properties
      % recommendation average popularity
      % recommendation diversity
      % spread (entropy of recommendations) (related to recommendation coverage)

  % dataset
    % # user
    % # items
    % distributions of # ratings per user (important: min 20)

  % temporal replay evaluation 
    % similar technique, harder to do than ours, and with a different goal.

  %TODO: restructure this, I can't have my cake and eat  too. This needs ot be refocused to explicitly focus on similarities and differences.
  %One related approach to the one we take here is to evaluate recommender systems temporally.
  %Various authors have taken this approach \cite{KorenTemporal, LathiaTemporal, burke2010evaluating} to evaluating recommender systems to understand how recommenders change with time.
  %This work often features similar evaluations to ours, focusing on time (life of user or life of system) as opposed to number of ratings.
  %One major difference between this approach and the one we take is that using ratings in the order that they were given can expose algorithms to biases from the system they were collected in.
  %For example, in the MovieLens system users enter their first 15 ratings in a different interface, which does not give them choice over which items to rate.
  %This is a forced situation which produces qualitatively different ratings than normal ratings.
  %For this work we want to avoid this effect and therefor must use randomization.
  %TODO: SOMEWHERE AROUND HERE THIS GOES!

  % subsampling strategy
    % describe normal crossfold solution
    % describe issues with solution
    % describe our crossfold then subsample solution.

@@

  Using the methodology described in section \ref{sec:methodology} we compute a value for each metric on each of the five algorithms.
  We plot these results against the number of items retained.
  These plots allow us to compare the algorithms against each other, and also, to understand how each algorithm's performance changes with more data.
  Each plot will have error bars showing the 95\% confidence intervals around the average per use value for the metric.

\subsection{Prediction accuracy}
  % Prediction accuracy
\begin{figure*}
  \centering

  \includegraphics[width=\linewidth]{../lenskit/output/ekstrandTuned20/accuracy.pdf}
  \caption{TODO}
  \label{fig:rmse}
  \label{fig:ndcg}
\end{figure*}
    % RMSE plot
    % point out error bars etc.
    % explain / interpret plot
  Figure \ref{fig:rmse} shows how accurate our algorithms are as measured by RMSE.
  As would be expected the accuracy of our item baseline doesn't change as the amount of information changes.
  The User-User and svd algorithms both perform as expected, starting out with the same accuracy as the item baseline, and then monotonically improving.
  User-User does show a small increase of error with only one rating, but the change is quite small and unlikely to be noticed by users.
  Interestingly, User-User and svd both perform about as well as our smarter user item baseline. %TODO: explain this better

@@

\begin{figure*}[ht!]
  \centering
  \includegraphics[width=1.1\linewidth]{../lenskit/output/ekstrandTuned20/TopNPrecision.pdf}
  \caption{TODO}
  \label{fig:map}
\end{figure*}

@@

  % TopN average rating / topn rmse
    % interpret graph and reason about what it means about quality

\begin{figure*}[ht!]
  \centering
  \includegraphics[width=1.1\linewidth]{../lenskit/output/ekstrandTuned20/rmse_20.pdf}
  \caption{TODO}
  \label{fig:topN.rmse}
\end{figure*}

  The first graph in figure \ref{fig:topN.rmse} shows the average number of items in the top 20 recommendations for each algorithm that were also in the test set.
  The average number of test set items in recommendations shows a similar trend to the MAP, precision, and recall metrics, user-user getting by far the fewest seen items, and the baseline getting the most.
  %TODO: talk about trust building here!

@@

\subsection{Other recommender metrics}
  % Other recommender metrics

\begin{figure*}[ht!]
  \centering
  \includegraphics[width=\linewidth]{../lenskit/output/ekstrandTuned20/popdiv.pdf}
  \caption{TODO}
  \label{fig:pop}
\end{figure*}

    % popularity graph
  The first graph in figure \ref{fig:pop} shows the average popularity of recommended items by each algorithm.
  %TODO: re-write this
  Across all datapoint sizes the ranking of algorithms is the same, with the baselines recommending the most popular movies, followed by svd, which recommends less popular movies, then Item-Item, which recommends much less popular movies, leaving User-User to recommend the least popular movies.
  Unfortunately, we don't know of any guideline for understanding how popular users desire their recommendations to be.
  That in mind, we expect that users will feel the baseline's popularity (around 1000) is too high, and User-User's popularity (around 50) is too low.
  SVD and Item-Item both seem to perform well, with SVD generating more popular recommendations, than Item-Item, and both algorithms showing a significant trend to make less popular predictions as we know more about the user.

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=1.1\columnwidth]{../lenskit/output/ekstrandTuned20/topN_entropy.pdf}
  \caption{TODO}
  \label{fig:spread}
\end{figure}
